An R Markdown document converted from "03_ValClusOutcomes.ipynb

### 2023-08-20 03_ValClusOutcomes.Rmd
Add files via upload

If any of the columns in this table are missing in your data, one option is to assume that none in your population had the disease, i.e., you should have a column with all 0.

@@

```{r}
alldat <- mutate(
    validclusmod,
    data = map(sex, ~strat_dat[[.x]])
)
print(alldat)
```

Once we have this table, we can run the function to calculate cluster probabilities:

```{r}
alldat <- clusterprobcalc(alldat)
print(alldat)
```

@@

```{r}
head(alldat$data[[1]])
```

@@

```{r}
markerdistribdf <- markerdistribfx(alldat)
```

@@

```{r}
bmieffmarkerdf <- bmieffmarkerfx(alldat)
```

@@

```{r}
alldat <- mutate(
    alldat,
    data = map(data, inner_join, covar_dat, by = "eid")
)
print(alldat)
```

@@

```{r}
countcovarsdf <- countcovarsfx(alldat)
```

@@

We are also interesting in looking at the following proportions in each cluster:

- Proportion of individuals with hypertension receiving antihypertensives.
- Proportion of individuals with CHD receiving lipid-lowering medication.
- Proportion of individuals with T2D taking insulin.
- Proportion of individuals with T2D taking insulin or any other anti-diabetic medication.

```{r}
countspectxdf <- countspectxfx(alldat)
```

```{r}
head(countspectxdf)
```

@@

```{r}
assocdxdf <- assocdxfx(alldat)
```

```{r}
head(assocdxdf)
```

@@

```{r}
alldat <- alldat %>%
    mutate(
        macedf = purrr::map(data, addsurvmacedat, SURVDATA = survmacedat),
        dmdf = purrr::map(data, addsurvdmdat, SURVDATA = survdmdat)
    )
print(alldat)
```

`macedf` and `dmdf` now contain the data necessary to run survival analysis.

@@

## Overall and cluster-specific Kaplan-Meier estimates

The first thing to do is to obtain overall and cluster-specific cumulative incidence rates using the Kaplan-Meier method:

```{r}
kmestdf <- kmestfx(alldat)
```

@@

### MACE

@@

@@

```{r}
coxmodmacedf <- coxmodelsmace(alldat)
```

```{r}
print(coxmodmacedf)
```

Here `score2` contains the predictors used in SCORE2-Diabetes, plus other predictors that we have in our input table and cluster allocations. The two models are contained in the last two columns. `mod_null` contains the null model, which we will use to calculate our metrics.

@@

```{r}
macesurvcoefs <- macesurvcoefx(coxmodmacedf)
```

```{r}
head(macesurvcoefs)
```

@@

We can now assess the predictive ability of each model and compare them. As the models are nested, we will use the gold-standard method, the likelihood ratio test. Given the wide use of the c-statistic, we will also use this metric. However, comparing two c-statistics is not as powerful as the likelihood ratio test.

```{r}
compmodmacedf <- comparemodsmace(coxmodmacedf)
compmodmacedf
```

@@

While p-values of both the likelihood ratio tests and the difference between C-statistics show evidence of added value, the adequacy index quantifies how much additional information is obtained when cluster allocations are added on top of common biomarkers contained the SCORE2.

@@

```{r}
adeqindbyclusmace <- AdeqIndClusMACEFx(coxmodmacedf)
```

```{r}
adeqindbyclusmace
```

#### Adequacy index by SCORE2 probabilities

We will also assess the adequacy index in individuals over certain thresholds of 10-year probability of MACE calculated by SCORE2. This is to have an idea of the utility of adding cluster allocation information across the scale of SCALE2.

```{r}
adeqindbypremace <- AdeqIndByPreMACEFx(coxmodmacedf)
```

```{r}
adeqindbypremace
```

@@

```{r}
dcamace <- DCurvMACEFx(coxmodmacedf)
head(dcamace)
```

@@

```{r}
dcaclusmace <- DCurvMACEbyClFx(coxmodmacedf)
```

```{r}
head(dcaclusmace)
```

### Diabetes

For diabetes progression, the reference model will include fasting glucose, in addition to all the components of the metabolic syndrome that we already have included in our input table. And as before, the second model will include cluster probabilities.

```{r}
coxmoddmdf <- coxmodelsdm(alldat)
```

```{r}
print(coxmoddmdf)
```

```{r}
dmsurvcoefs <- dmsurvcoefx(coxmoddmdf)
```

```{r}
head(dmsurvcoefs)
```

```{r}
compmoddmdf <- comparemodsdm(coxmoddmdf)
compmoddmdf
```

```{r}
adeqindbyclusdm <- AdeqIndClusDMFx(coxmoddmdf)
```

```{r}
adeqindbyclusdm
```

#### Adequacy index by probabilities of base model

```{r}
adeqindbypredm <- AdeqIndByPreDMFx(coxmoddmdf)
```

```{r}
adeqindbypredm
```

```{r}
dcadm <- DCurvDMFx(coxmoddmdf)
head(dcadm)
```

```{r}
dcaclusdm <- DCurvDMbyClFx(coxmoddmdf)
```

```{r}
head(dcaclusdm)
```

@@

```{r}
result_file2 <- list(
    MarkerDistrib = markerdistribdf,
    BMIeffOnMarker = bmieffmarkerdf,
    CountCovars = countcovarsdf,
    CountSpecDXMeds = countspectxdf,
    CrossSectAssoc = assocdxdf,
    KaplanMeierDF = kmestdf,
    MACESurvCoefs = macesurvcoefs,
    ComparisonMACE = compmodmacedf,
    AdeqIndByClusMACE = adeqindbyclusmace,
    AdeqIndByPreMACE = adeqindbypremace,
    DCAResMACE = dcamace,
    DCAREsClusMACE = dcaclusmace,
    DMSurvCoefs = dmsurvcoefs,
    ComparisonDM = compmoddmdf,
    AdeqIndByClusDM = adeqindbyclusdm,
    AdeqIndByPreDM = adeqindbypredm,
    DCAResDM = dcadm,
    DCAREsClusDM = dcaclusdm
)
```

@@

<!-- BIBLIOGRAPHY START -->
<div class="csl-bib-body">
  <div class="csl-entry"><i id="zotero|10831815/ZY2CL5NC"></i>SCORE2 working group and ESC Cardiovascular risk collaboration. (2021). SCORE2 risk prediction algorithms: new models to estimate 10-year risk of cardiovascular disease in Europe. <i>European Heart Journal</i>, <i>42</i>(25), 2439–2454. https://doi.org/10.1093/eurheartj/ehab309</div>
  <div class="csl-entry"><i id="zotero|10831815/FPUFQKFI"></i>SCORE2-Diabetes Working Group and the ESC Cardiovascular Risk Collaboration. (2023). SCORE2-Diabetes: 10-year cardiovascular risk estimation in type 2 diabetes in Europe. <i>European Heart Journal</i>, ehad260. https://doi.org/10.1093/eurheartj/ehad260</div>
</div>
<!-- BIBLIOGRAPHY END -->

>>>>

If any of the columns in this table are missing in your data, one option is to assume that none in your population had the disease, i.e., you should have a column with 0 for all individuals.

@@

```{r}
clusterdfs <- clusterprobcalc(ClusModDf = validclusmod, StratDat = strat_dat)
```

```{r}
print(clusterdfs)
```

@@

```{r}
head(clusterdfs$data[[1]])
```

@@

```{r}
markerdistribdf <- markerdistribfx(clusterdfs)
```

@@

```{r}
bmieffmarkerdf <- bmieffmarkerfx(clusterdfs)
```

@@

```{r}
clusterdfs <- addcovardat(X = clusterdfs, CovarDat = covar_dat)
```

```{r}
print(clusterdfs)
```

@@

```{r}
countcovarsdf <- countcovarsfx(clusterdfs)
```

@@

We are also interesting in looking at the proportion of individuals receiving medications in each cluster, stratified by each condition. This is obtained with the following function:

```{r}
countdxmed <- countdxmedfx(clusterdfs)
```

```{r}
head(countdxmed)
```

@@

```{r}
assocdxdf <- assocdxfx(clusterdfs)
```

```{r}
print(assocdxdf)
```

@@

```{r}
clustersurvdfs <- addsurvdat(X = clusterdfs, SurvMACEDf = survmacedat, SurvDMDf = survdmdat)
```

```{r}
print(clustersurvdfs)
```

`data` now contain the data necessary to run survival analysis.

@@

## Rates of outcomes by cluster

Similar to what was done in the cross sectiona setting, we will calculate the number of cases and the total follow-up in each cluster using the weighted approach:

```{r}
ratesbyclus <- ratesclusfx(clustersurvdfs)
```

```{r}
head(ratesbyclus)
```

We will also do this by medication status:

```{r}
ratesbyclusmed <- ratesclusmedfx(clustersurvdfs)
```

```{r}
head(ratesbyclusmed)
```

We will also calculate overall and cluster-specific cumulative incidence rates using the Kaplan-Meier method:

```{r}
kmestdf <- kmestfx(clustersurvdfs)
```

@@

@@

Similarly, for diabetes we will also fit two models, one containing all biomarkers and another one containing the biomarkers plus the cluster probabilities.

The way we will introduce the cluster probabilities into the Cox models will be employing the log-ratio transformation <cite id="0lvet"><a href="#zotero|10831815/UYLBIXH3">(Coenders &#38; Pawlowsky-Glahn, 2020)</a></cite>.

@@

```{r}
coxmoddf <- coxmodels(clustersurvdfs)
```

```{r}
print(coxmoddf)
```

Here `mod_base` contains the baseline model, while `mod_clus` contains the baseline plus clusters model. `mod_null` contains the null model, which we will use to calculate our metrics.

@@

To properly calculate the effect of clusters, we need the coefficients estimated by the log-ratio models as well as their covariance, contained in the `estimates` and `varcovmat` columns. To properly calculate the expected risk for a given phenotype, we need first the means of all variables in the model, contained in the column `Means` and the parameters of the baseline hazard, contained in the `Afit` column.

```{r}
survcoefs <- survcoefx(coxmoddf)
```

```{r}
print(survcoefs)
```

@@

To assess the predictive ability of the two nested models, we will use the gold-standard method: the likelihood ratio test. Given the wide use of the c-statistic, we will also use this metric. However, comparing two c-statistics is not as powerful as the likelihood ratio test.

```{r}
compmoddf <- comparemods(coxmoddf)
```

```{r}
compmoddf
```

@@

While p-values of both the likelihood ratio tests and the difference between C-statistics show evidence of added value, the adequacy index quantifies how much additional information is obtained when cluster allocations are added on top of the baseline model.

@@

```{r}
adeqindbyclus <- AdeqIndClusFx(coxmoddf)
```

```{r}
adeqindbyclus
```

#### Adequacy index by MACE probability given by baseline model

We are also interested in how the more complex model behaves along the scale of MACE probabilities given by the baseline model:

```{r}
adeqindbypre <- AdeqIndByPreFx(coxmoddf)
```

```{r}
head(adeqindbypre)
```

@@

```{r}
dcares <- DCurvFx(coxmoddf)
```

```{r}
head(dcares)
```

@@

```{r}
dcaclusres <- DCurvbyClFx(coxmoddf)
```

```{r}
head(dcaclusres)
```

## Interaction between clusters and medications

Finally, we will assess the interaction between certain medications and clusters:

```{r}
interactmods <- interactmodfx(coxmoddf)
```

```{r}
print(interactmods)
```

@@

```{r}
result_file2 <- list(
    MarkerDistrib = markerdistribdf,
    BMIeffOnMarker = bmieffmarkerdf,
    CountCovars = countcovarsdf,
    CountDXMeds = countdxmed,
    CrossSectAssoc = assocdxdf,
    RatesByClus = ratesbyclus,
    RatesByClusMeds = ratesbyclusmed,
    KaplanMeierDF = kmestdf,
    SurvCoefs = survcoefs,
    Comparison = compmoddf,
    AdeqIndByClus = adeqindbyclus,
    AdeqIndByPre = adeqindbypre,
    DCARes = dcares,
    DCAClusREs = dcaclusres,
    InteractMods = interactmods
)
```

@@

<!-- BIBLIOGRAPHY START -->
<div class="csl-bib-body">
  <div class="csl-entry"><i id="zotero|10831815/UYLBIXH3"></i>Coenders, G., &#38; Pawlowsky-Glahn, V. (2020). On interpretations of tests and effect sizes in regression models with a compositional predictor. <i>SORT-Statistics and Operations Research Transactions</i>, <i>44</i>(1), 201–220. https://doi.org/10.2436/20.8080.02.100</div>
  <div class="csl-entry"><i id="zotero|10831815/ZY2CL5NC"></i>SCORE2 working group and ESC Cardiovascular risk collaboration. (2021). SCORE2 risk prediction algorithms: new models to estimate 10-year risk of cardiovascular disease in Europe. <i>European Heart Journal</i>, <i>42</i>(25), 2439–2454. https://doi.org/10.1093/eurheartj/ehab309</div>
  <div class="csl-entry"><i id="zotero|10831815/FPUFQKFI"></i>SCORE2-Diabetes Working Group and the ESC Cardiovascular Risk Collaboration. (2023). SCORE2-Diabetes: 10-year cardiovascular risk estimation in type 2 diabetes in Europe. <i>European Heart Journal</i>, ehad260. https://doi.org/10.1093/eurheartj/ehad260</div>
</div>
<!-- BIBLIOGRAPHY END -->
